---
title: "Data Carpentry Workshop Remix Jan 15, 2020"
author: "Rachel Schwartz"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

Check out the [Data Carpentry Intro](https://datacarpentry.org/R-ecology-lesson/00-before-we-start.html) before starting here.

Note that this lesson is adapted from the [Data Carpentry Ecology Lesson](https://datacarpentry.org/R-ecology-lesson/index.html). You can find a lot more information there. Text that has been copied from this lesson is highlighted in blue.

```{r pack, warning=FALSE}
library(tidyverse)
```

## Get the Data

The data used for this less is all about Australia, including its climate over time and recent fires. To learn more about Australia's devastating fires check out this NY Times [article](https://www.nytimes.com/interactive/2020/01/02/climate/australia-fires-map.html). The data and information comes from Tidy Tuesday, a weekly data project aimed at the R ecosystem. You can find more datasets at https://github.com/rfordatascience/tidytuesday . 

The data are available online so the first thing we'll do is load them into R.
To load data you'll need to use a *function*.

<p style="color:blue">
Functions are "canned scripts" that automate more complicated sets of commands
including operations assignments, etc. Many functions are predefined, or can be
made available by importing R *packages* (more on that later). A function
usually takes one or more inputs called *arguments*. Functions often (but not
always) return a *value*. A typical example would be the function `sqrt()`. The
input (the argument) must be a number, and the return value (in fact, the
output) is the square root of that number. Executing a function ('running it')
is called *calling* the function.
</p>

We'll read in our data using the `read_csv()` function, from the tidyverse package **`readr`**.
We assign each dataset to a variable so we can reuse it.
The name of the variable is on the left side of the arrow.
The action we're doing with the function is on the write side.

```{r data}
rainfall <- read_csv('https://tinyurl.com/Oz-fire-rain')
temperature <- read_csv('https://tinyurl.com/Oz-fire-temp')
```

You can see some information about the data we have just loaded.
The name of each column is shown along with the type of data in that column.
The data are stored in a format we call a data frame.
For more details on this dataset see the [Tidy Tuesday site](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-01-07).

<p style="color:blue">
This statement doesn’t produce any output because assignments don’t display anything. 
If we want to check that our data has been loaded, we can see the contents of the data frame by typing its name:`rainfall`.
</p>

```{r d2}
rainfall
```

You can also view the data in a separate window by clicking on its name in the Global Environment window.
Here you can also see some information about the data. Note how much data we have!

Note that we've read our data from a website, but you can read local files as well.
These files are in csv format, which means plain text where columns are separated by commas.
This is a very simple format that avoids all the complexity of Excel.
If you need to read an Excel file you can either export as a csv or use a different function to read the data.

For more information on data frames see the [Starting with Data](https://datacarpentry.org/R-ecology-lesson/02-starting-with-data.html#what_are_data_frames) section of the Data Carpentry lesson.

## Basic data exploration

For a really basic exploratory analysis let's look at how temperature is changing in Australian cities over time.
I've summarized our original dataset so you can make your first plot more easily.
First: read your data from 

```{r t}

ggplot(rainfall, aes(month,rainfall)) +
  geom_point()
```
